Individual Variation in Animal Movement and Effects on Seed Dispersal Distance

### 2020-12-28 paper/paper.Rmd
update article

---
title: "Individual Variation in Animal Movement and Effects on Seed Dispersal Distance"

abstract: |
  Seed dispersal is a fundamental process in plant ecology and due to its high importance in the ecosystem, various modelling approaches have been developed to understand seed movement and dispersal mechanisms. In particular, long-distance seed dispersal (LDD) plays a significant role in determining a population's genetic diversity and range expansion. Animal-mediated seed dispersal is a key component of LDD events, as frugivores can travel long distances and transport seeds with them. In this study we seek to understand the implications of individual variation in animal movement and how these can impact estimates of long distance seed dispersal. We apply these methods to empirical data sets that contain locations for an aracari, *Pteroglossus pluricintus*, one of the primary frugivores for the Amazonian canopy tree *Virola flexuosa*. We combine animal movement data and gut retention time to simulate *Virola* seed dispersal. This approach showed that there is significant variation in individual aracari movement patterns which directly influence the estimated number of long-distance dispersal events for *Virola* seeds. We found that models that ignore this underlying variation in individual animal movement, underestimate the percentage of long-distance seed dispersal events and maximum seed dispersal distances.
output: word_document
bibliography: mybibfile.bib
---

@@

In order to analyze the effect of differences in animal movement over final dispersal distances of foraged seeds we developed a spatially-explicit individual-based model in a homogeneous landscape that incorporated animal movement and seed gut retention time (GRT). Our landscape consisted of a single source tree at the origin, our simulation time steps were based on one minute intervals, and each simulation started at the source tree (0,0), where each individual bird was given five seeds and was allowed to move freely in the landscape until the simulation ended once all seeds were dropped. Each seed's gut retention time was determined at the start of each simulation run, being randomly sampled from a Gamma distribution (shape = 4, scale = 5, shift = 8) [@morales_effects_2006] so that the average gut retention time was 28 minutes[@holbrook_dispersal_2009; @holbrook_using_2007]. For each simulation run, the total simulation time was determined by the largest gut retention time sampled for the five seeds, therefore, once the last seed was dropped, the simulation ended. From the simulation, we calculated the Euclidean distance from the source tree to each of the dropped seeds to determine the seed dispersal distance. We simulated animal movement using a memoryless process, a random walk with a uniformly distributed angular direction and a random step size chosen from a probability distribution.

@@

Our main hypothesis is about the incoporation of variance in animal movement at different organizational levels, meaning how are the differences in individual animal movement affecting population level estimates of seed dispersal. How does the underlying variation in individual animal movement impact our estimates of long distance seed dispersal. We explored the effects of individual differences in movement by modifying our base simulation, described above, in which the probability density function used to sample step sizes was different between population, individual, or family level simulations. What we consider as population level simulations are those where all birds share the same probability density to sample step sizes from, while for individual and family level simulations we use the same overall pdf, but allow for variation in the parameters specified for such function, where each set of parameters corresponds to a specific individual or family group. We also performed simulations in which we allowed for variation not only of parameters, but also of the overall structure of the pdf for sampling step sizes, this means that each individual or family group was allowed to have it's own pdf and its own set of parameters (Box 1 or table 1 with description of models, either the movement or simulation). We used four different probability density functions for sampling the animal movement step size, these included an exponential distribution [@jones_closing_2017], gamma distribution [citation], Weibull distribution[citation], and lognormal distribution [@levey_modelling_2008]. These four distributions are leptokurtic, and they vary in the fatness of their tails, thus allowing for different levels of variation in step size. **should add number of runs? or mention that that is in appendix B**

@@

We focused our study on the many-banded aracari (*Pterglossus pluricinctus*) and its role as a frugivore of the *Virola flexuosa* tree. A subset of animal movement data collected in a previous study [@holbrook_home_2011] was selected and used towards our models and simulations. In this previous study, Aracari home ranges, average movement distances, patterns and movement rates were estimated for tracked birds over a period of four years, from 2001 to 2005, in the Ecuadorian Amazon rainforest. Further details on field methods can be found in Holbrook [-@holbrook_home_2011]. We focused on the data from the many-banded aracari, *Pteroglossus pluricinctus*, because a higher number of recorded locations were available. In addition to this, the home range of this species was found to be the largest for all the toucan species studied, and therefore it was better suited to study long-distance seed dispersal events.

>>>>

---
title: "Individual Variation in Animal Movement and Effects on Seed Dispersal Distance"
abstract: |
  Seed dispersal is a fundamental process in plant ecology and due to its high importance in the ecosystem, various modelling approaches have been developed to understand seed movement and dispersal mechanisms. In particular, long-distance seed dispersal (LDD) plays a significant role in determining a population's genetic diversity and range expansion. Animal-mediated seed dispersal is a key component of LDD events, as frugivores can travel long distances and transport seeds with them. In this study we seek to understand the implications of individual variation in animal movement and how these can impact estimates of long distance seed dispersal. We apply these methods to empirical data sets that contain locations for an araçari, *Pteroglossus pluricintus*, one of the primary frugivores for the Amazonian canopy tree *Virola flexuosa*. We combine animal movement data and gut retention time to simulate *Virola* seed dispersal. This approach showed that there is significant variation in individual araçari movement patterns which directly influence the estimated number of long-distance dispersal events for *Virola* seeds. We found that models that ignore this underlying variation in individual animal movement, underestimate the percentage of long-distance seed dispersal events and maximum seed dispersal distances.
output:
  word_document: default
  html_document:
    df_print: paged
bibliography: mybibfile.bib
---

@@

In order to analyze the effect of differences in animal movement over final dispersal distances of foraged seeds we developed a spatially-explicit individual-based model in a homogeneous landscape that incorporated animal movement and seed gut retention time (GRT). Our landscape consisted of a single source tree at the origin, our simulation time steps were based on one minute intervals, and each simulation started at the source tree (0,0), where each individual bird was given five seeds and was allowed to move freely in the landscape until the simulation ended once all seeds were dropped, Figure1. Each seed's gut retention time was determined at the start of each simulation run, being randomly sampled from a Gamma distribution (shape = 4, scale = 5, shift = 8) [@morales_effects_2006] so that the average gut retention time was 28 minutes[@holbrook_dispersal_2009; @holbrook_using_2007]. For each simulation run, the total simulation time was determined by the largest gut retention time sampled for the five seeds, therefore, once the last seed was dropped, the simulation ended. We calculated the Euclidean distance from the source tree to each of the dropped seeds to determine their seed dispersal distance. We simulated animal movement using a memoryless process, a random walk with a uniformly distributed angular direction and a random step size chosen from a probability distribution. The probability distribution chosen to simulate step size varied according to the different models we propose in Table1.

@@

Our main hypothesis is about the incorporation of variation in animal movement at different organizational levels, meaning how are the differences in individual animal movement affecting population level estimates of seed dispersal. How does the underlying variation in individual animal movement impact our estimates of long distance seed dispersal. We explored the effects of individual differences in movement by modifying our base simulation, described above, in which the probability density function used to sample step sizes was different between population, individual, or family level simulations. What we consider as population level simulations are those where all birds share the same probability density to sample step sizes from, while for individual and family level simulations we use the same overall pdf, but allow for variation in the parameters specified for such function, where each set of parameters corresponds to a specific individual or family group. We also performed simulations in which we allowed for variation not only of parameters, but also of the overall structure of the pdf for sampling step sizes, this means that each individual or family group was allowed to have it's own pdf and its own set of parameters (Box 1 or table 1 with description of models, either the movement or simulation). We used four different probability density functions for sampling the animal movement step size, these included an exponential distribution [@jones_closing_2017], gamma distribution [citation], Weibull distribution[citation], and lognormal distribution [@levey_modelling_2008]. These four distributions are leptokurtic, and they vary in the fatness of their tails, thus allowing for different levels of variation in step size. **should add number of runs? or mention that that is in appendix B**

@@

We focused our study on the many-banded araçari (*Pterglossus pluricinctus*), a small toucan, and its role as a frugivore of the *Virola flexuosa* tree. A subset of animal movement data collected in a previous study [@holbrook_home_2011] was selected and used towards our models and simulations. In this previous study, araçari home ranges, average movement distances, patterns and movement rates were estimated for tracked birds over a period of four years, from 2001 to 2005, in the Ecuadorian Amazon rainforest. Further details on field methods can be found in Holbrook [-@holbrook_home_2011]. Although this previous study included two larger Ramphastids as well, we used data only from the many-banded araçari, because a higher number of recorded locations were available. In addition to this, the home range of this species was found to be the largest for all the species studied, and therefore it was better suited to study long-distance seed dispersal events.
